---
title: "Project for Week 4"
author: "Ryan Ordonez"
date: "2023-10-08"
output:
  html_document: default
  pdf_document: default
---

**This project is to identify certain boroughs of New York to impliment after school community centers to help reduce crime in the most incident prone areas for our youth.**  

### Load Packages
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(reshape2)
library(pROC)
```

### Import Data
```{r}
# breaking down the url so it doesn't run off the page of the pdf doc
url_part1 <- "https://data.cityofnewyork.us/api/views/833y-fsy8/"
url_part2 <- "rows.csv?accessType=DOWNLOAD"
NYPD_shootings <- read.csv(paste0(url_part1, url_part2))
```

**This data is pulled from the Data.gov website. It lists every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. A link to the .csv file is here:** <https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD>

### Inspect the Data
```{r}
head(NYPD_shootings)
```

### Tidy and Transform the data
```{r}
# Convert appropriate variables to factor types
NYPD_shootings$BORO = as.factor(NYPD_shootings$BORO)
NYPD_shootings$LOC_OF_OCCUR_DESC = as.factor(NYPD_shootings$LOC_OF_OCCUR_DESC)
NYPD_shootings$PRECINCT = as.factor(NYPD_shootings$PRECINCT)
NYPD_shootings$JURISDICTION_CODE = as.factor(NYPD_shootings$JURISDICTION_CODE)
NYPD_shootings$LOC_CLASSFCTN_DESC = as.factor(NYPD_shootings$LOC_CLASSFCTN_DESC)
NYPD_shootings$STATISTICAL_MURDER_FLAG = as.factor(NYPD_shootings$STATISTICAL_MURDER_FLAG)
NYPD_shootings$PERP_AGE_GROUP = as.factor(NYPD_shootings$PERP_AGE_GROUP)
NYPD_shootings$PERP_SEX = as.factor(NYPD_shootings$PERP_SEX)
NYPD_shootings$PERP_RACE = as.factor(NYPD_shootings$PERP_RACE)
NYPD_shootings$VIC_AGE_GROUP = as.factor(NYPD_shootings$VIC_AGE_GROUP)
NYPD_shootings$VIC_SEX = as.factor(NYPD_shootings$VIC_SEX)
NYPD_shootings$VIC_RACE = as.factor(NYPD_shootings$VIC_RACE)
# Remove unnecessary columns
NYPD_shootings$Lon_Lat = NULL
NYPD_shootings$X_COORD_CD = NULL
NYPD_shootings$Y_COORD_CD = NULL
NYPD_shootings$JURISDICTION_CODE = NULL
NYPD_shootings$LOC_CLASSFCTN_DESC = NULL
NYPD_shootings$LOCATION_DESC = NULL
NYPD_shootings$STATISTICAL_MURDER_FLAG = NULL
NYPD_shootings$Latitude = NULL
NYPD_shootings$Longitude = NULL
NYPD_shootings$LOC_OF_OCCUR_DESC = NULL
```

```{r}
# Mark null, unknown and empty entries as NA for easy processing
NYPD_shootings$PERP_AGE_GROUP[NYPD_shootings$PERP_AGE_GROUP %in% 
                                c("", "UNKNOWN", "(null)")] <- NA
```

```{r}
# Keep only specific age groups
valid_age_groups <- c("<18", "18-24", "25-44", "45-64", "65+", NA)
NYPD_shootings <- NYPD_shootings[NYPD_shootings$PERP_AGE_GROUP %in% valid_age_groups, ]
```

```{r}
summary(NYPD_shootings)
```

### Visual Analysis
```{r warning=FALSE}
# Filter out NA values from the dataset
NYPD_shootings_filtered <- NYPD_shootings %>% filter(!is.na(PERP_AGE_GROUP))

# Wrap the BORO column text for better display
NYPD_shootings_filtered$BORO <- str_wrap(NYPD_shootings_filtered$BORO, width = 10)

# Create a bar chart
ggplot(NYPD_shootings_filtered, aes(x = PERP_AGE_GROUP)) +
  geom_bar(aes(fill = PERP_AGE_GROUP), position = "dodge") +
  ggtitle("Number of Incidents by Age Group and Boro") +
  xlab("Age Group") +
  ylab("Number of Incidents") +
  facet_wrap(~ BORO, ncol = length(unique(NYPD_shootings_filtered$BORO))) +
  theme(
    axis.text.x = element_blank(),  # Remove x-axis text
    axis.ticks.x = element_blank()  # Remove x-axis ticks
  )
```

**The above chart identifies that most incidents occur in the Bronx and Brooklyn boroughs by perpetrators in the two age groups of 18-24 and 25-44. I will now dive into the hours in which most of these incidents occur within these boroughs.** 

### Time-Based Analysis for Bronx and Brooklyn
```{r}
# Filter data for Bronx and Brooklyn
NYPD_Bronx_Brooklyn <- NYPD_shootings %>% filter(BORO %in% c("BRONX", "BROOKLYN"))
```

```{r}
# Extract hour information (this assumes 'TIME_OCCURED' is a POSIX time object)
NYPD_Bronx_Brooklyn$HOUR_OCCURED <- format(as.POSIXct(NYPD_Bronx_Brooklyn$OCCUR_TIME, format="%H:%M:%S"), "%H")
```

```{r}
# Filter the data for the age groups 18-24 and 25-44
filtered_data <- NYPD_Bronx_Brooklyn %>% 
  filter(PERP_AGE_GROUP %in% c('18-24', '25-44'))

# Prepare the title with wrapped text
wrapped_title <- strwrap("Number of Incidents by Hour in Bronx and Brooklyn commited by perps of ages 18-44", 
                         width = 30) %>%
  paste(collapse = "\n")

# Create a bar chart for incidents by time of day in Bronx and Brooklyn
ggplot(filtered_data, aes(x = HOUR_OCCURED)) +
  geom_bar(aes(fill = HOUR_OCCURED), position = "dodge") +
  ggtitle(wrapped_title) +
  theme(plot.title = element_text(hjust = 0.5)) +  # Center the title
  xlab("Hour of the Day") +
  ylab("Number of Incidents")
```


**This chart shows that incidents by these two age groups in these two boroughs occur mostly from 1400 (2pm) to 0400 (4am). An after school facility in these areas may help reduce the amount of crimes committed in these areas between these times. Some questions to take into consideration is what hours should the facility be open, what type of activities would it provide to reduce crime, how can we promote the facility, and how can we provide transportation to and from the facility for our youth?.**   


### Modeling
```{r}
selected_data <- filtered_data %>%
  select(HOUR_OCCURED, BORO, PERP_AGE_GROUP)

selected_data$HOUR_OCCURED <- as.numeric(selected_data$HOUR_OCCURED)
selected_data$BORO <- as.factor(selected_data$BORO)
selected_data$PERP_AGE_GROUP <- as.factor(selected_data$PERP_AGE_GROUP)

# Linear regression
model <- glm(HOUR_OCCURED ~ BORO + PERP_AGE_GROUP, data = selected_data, family = "gaussian")

summary(model)
```

Based on the GLM results, one key finding is the significant negative coefficient for the perpetrator age group 25-44, which suggests that crimes committed by individuals in this older age group occur, on average, 0.86 hours earlier than those in the reference age group. This could be interpreted as older individuals being involved in crime earlier in the day, leaving room to infer that younger individuals (such as ages 18-24, if they are the reference group) are more likely to be involved in crimes that occur later in the day.

Moreover, while the coefficient for the Brooklyn borough is not statistically significant, it does indicate a positive effect on the hour crimes occur, suggesting crimes happen slightly later in Brooklyn compared to the reference borough. When combining these insights, it's reasonable to propose that after-school activities targeting youth ages 18-24 in Brooklyn could potentially fill the time gap where they're more likely to engage in criminal activities, thus reducing crime rates in this particular demographic and area.

### Conclusion
The data clearly shows trends in the age groups of the perpetrators. Specifically, the 18-24 year-old group leads in the number of incidents, and this is most noticeable in Brooklyn borough.

Some questions that arise from this data include the differences in policing policies in each borough. There's also a concern about the large number of unreported or missing age groups that could potentially skew our understanding of incidents by age group.

In terms of bias, how data is reported can differ from precinct to precinct. Additionally, the unique policies of each precinct could introduce underlying biases, especially when incidents by age group have historically been treated differently.

On a personal note, I place a high degree of trust in data. While I have tried to address missing or erroneous entries, there's always the risk of the data being off or misleading. To mitigate this, I've focused on maintaining the integrity of the majority of correct data while considering the impact of erroneous entries. I also plan on mitigating my own bias towards quantitative data by corroborating these findings with other data sources and incorporating them into a single, comprehensive report.